---
title: "PCA_tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(root.dir = rprojroot::find_rstudio_root_file(), collapse = TRUE, comment = "#>") # Set working directory of Rmd to package root directory
```

```{r setup}
library(glab.library2)
```

# PCA / PLSR Tutorial

Originally written by Katherine Sheu.
Recommend running this tutorial using RStudio.

## Introduction

Here are some examples for running PCA and PLSR using functions from this package. The tutorial is written in R Markdown format -- similar to Jupyter Notebooks for Python where code chunks are interspersed between descriptive text and can be run.
  
## PCA Run & Plot

Scripts:
glab.library2/R/
  PCA_from_file.R
	plot_PCA.R

The PCA scripts are basically a wrapper for the R package 'prcomp'.
	https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html

Input data:
glab.library2/inst/extdata/
  Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt
	human.info.rsem.expression.txt

As the file name for the data implies, the input data comes from a 2016 paper from Beltran et al. Citation:
Beltran H, Prandi D, Mosquera JM, Benelli M, Puca L, Cyrta J, Marotz C, Giannopoulou E, Chakravarthi BV, Varambally S, Tomlins SA, Nanus DM, Tagawa ST, Van Allen EM, Elemento O, Sboner A, Garraway LA, Rubin MA, Demichelis F. Divergent clonal evolution of castration-resistant neuroendocrine prostate cancer. Nat Med. 2016 Mar;22(3):298-305. doi: 10.1038/nm.4045. Epub 2016 Feb 8. PMID: 26855148; PMCID: PMC4777652.

Source the PCA R files to load the functions into memory.

```{r Source PCA functions}
source("../R/PCA_from_file.R")
source("../R/plot_PCA.R")
```

Functions that are installed as part of an R package such as this one (glab.library2) can also be called in code directly from the package as shown below. Use '?' before any function to view R Documentation on the Help pane.

```{r Call function from package}
?glab.library2::PCA_from_file()
?glab.library2::plot_PCA()
```

Run PCA using the function in this package. The function will output 3 text files associated with the PCA run to the folder that the gene expression counts are located (../inst/extdata/):
- Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_loadings.txt
- Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt
- Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_sdev.txt
```{r Run PCA from file}
PCA_from_file("../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt", center = T, scale = F)
```

Load in the annotation file for the samples. This file contains a table matching each sample with its type. The annotations are then used to color the points in the PCA plot of scores (the samples) by their type (LUAD, ESCA, etc.)
```{r Annotate plot by type, fig.width=7, fig.height=5}
human.info <- read.delim("../inst/extdata/human.info.rsem.expression.txt")

plot_PCA(file = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt",
         info.name = human.info$sample,
         info.type = human.info$type,
         labels = F, ellipse = T, density = T,
         title = "Beltran 2016 PCA Scores",
         PCx = "PC1", PCy = "PC2")
```



## PLSR Run & Plot

Scripts:
glab.library2/R/
	varimax_from_file.R
	PLSR_from_file.R
	plot_pls.R

Input files:
glab.library2/inst/extdata/
	Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt
	human.info.rsem.expression.txt


PLSR scripts are a wrapper for R package 'pls'mixOmics'.
  https://cran.r-project.org/web/packages/mixOmics/index.html
  http://www.bioconductor.org/packages/release/bioc/html/mixOmics.html
