---
title: "PCA / PLSR Tutorial"
author: "Katherine Sheu, Yi Jou (Ruby) Liao"
output: rmarkdown::html_vignette
fig_width: 7
fig_height: 5
vignette: >
  %\VignetteIndexEntry{PCA_tutorial}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(root.dir = dirname(dirname(rstudioapi::getSourceEditorContext()$path)), collapse = TRUE, comment = "#>") # Set working directory of Rmd to package root directory
options(rmarkdown.html_vignette.check_title = FALSE)
library(glab.library2)
```

Originally written by Katherine Sheu. Recommend running this tutorial using RStudio.

## Introduction

Here are some examples for running PCA and PLSR using functions from this package. The tutorial is written in R Markdown format -- similar to Jupyter Notebooks for Python where code chunks are interspersed between descriptive text and can be run.

## PCA Run & Plot

Input data:

-   glab.library2/inst/extdata/
    -   Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt
    -   human.info.rsem.expression.txt

Scripts:

-   glab.library2/R/
    -   PCA_from_file.R
    -   plot_PCA.R

The PCA scripts are basically a wrapper for the R package 'prcomp'.

-   <https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html>

As the file name for the data implies, the input data comes from a 2016 paper from @beltran2016 .

Source the PCA R files to load the functions into memory.

```{r Source PCA functions}
source('../R/PCA_from_file.R')
source("../R/plot_PCA.R")
```

Functions that are installed as part of an R package such as this one (glab.library2) can also be called in code directly from the package as shown below. Use '?' before any function to view R Documentation on the Help pane.

```{r Call function from package}
?glab.library2::PCA_from_file()
?glab.library2::plot_PCA()
```

Run PCA using the function in this package. The function will output 3 text files associated with the PCA run to the folder that the gene expression counts are located (../inst/extdata/):

-   Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_loadings.txt
-   Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt
-   Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_sdev.txt

```{r Run PCA from file}
PCA_from_file("../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt", center = T, scale = F)
```

Load in the annotation file for the samples. This file contains a table matching each sample with its type. The annotations are then used to color the points in the PCA plot of scores (the samples) by their type (LUAD, ESCA, etc.)

```{r Load annotation file}
human.info <- read.delim("../inst/extdata/human.info.rsem.expression.txt")
head(human.info)
```

```{r Plot PCA with annotation info, fig.width=7, fig.height=5}
require(ggplot2)
plot_PCA(file = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt",
         info.name = human.info$sample,
         info.type = human.info$type,
         labels = F, ellipse = T, density = T,
         title = "Beltran 2016 PCA Scores",
         PCx = "PC1", PCy = "PC2")
```

## PLSR Run & Plot

Scripts:

-   glab.library2/R/
    -   varimax_from_file.R
    -   PLSR_from_file.R
    -   plot_pls.R

Input files:

-   glab.library2/inst/extdata/
    -   Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt
    -   human.info.rsem.expression.txt

PLSR scripts are a wrapper for R package 'pls'mixOmics':

- <https://cran.r-project.org/web/packages/mixOmics/index.html>
- <http://www.bioconductor.org/packages/release/bioc/html/mixOmics.html>


PLSR is a supervised algorithm, so the labels for samples (human.info) must be provided to indicate which samples are the response values. The if statement in function call assigns a "1" for samples with type == "NEPC" while all other types are assigned "0".
```{r Run PLSR from file}
# source('../R/PLSR_from_file.R')
human.info <- read.delim("../inst/extdata/human.info.rsem.expression.txt")

glab.library2::PLSR_from_file(
  file = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt",
  sample.names = human.info$sample,
  sample.type = human.info$type,
  y.response = (ifelse(human.info$type=="NEPC", 1, 0)),
  comps = 5, scale = F)
```


```{r Plot PLSR, fig.width=7, fig.height=5}
source('../R/plot_PLS.R')
human.info <- read.delim("../inst/extdata/human.info.rsem.expression.txt")

plot_PLS(
  file ="../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_PLSR_Xscores.txt",
  info.name = human.info$sample,
  info.type = human.info$type,
  title = "PLSR",
  labels = T,
  PCx = "comp1", PCy = "comp2", ellipse = T, conf = 0.90)
```

## Varimax

Varimax is a rotational transformation method that can be applied to either PCA or PLSR results to help increase interpretation of weights.

```{r Run varimax from file}
varimax_from_file(
  file.scores = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt",
  file.loadings = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_loadings.txt",
  comp = 2, normalize = F
)

# Plotting setup is the same as the previous PCA run
plot_PCA(
  file = "../inst/extdata/Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores_VARIMAX.txt",
  info.name = human.info$sample,
  info.type = human.info$type,
  labels = F, ellipse = T
)
```


